Ground state energy of quantum dots using the coupled cluster method

### 2018-04-16 project_2/article/article.tex
Add section on CCD energy

@@

\section{Theory}
    In this project we will study a system of $N$ interacting electrons. We will
    be looking at a Hamiltonian consisting of a one-body and a two-body part.
    The one-body part is given by
    \begin{align}
        h(\vf{r}_i)
        &= -\half\nabla_i^2 + \half\omega^2 \vf{r}_i^2,
    \end{align}
    where we use natural units $\hbar = c = e = 1$ and set the mass to unity.
    The two-body part is the Coulomb interaction potential.
    \begin{align}
        v(\vf{r}_i, \vf{r}_j)
        &= \frac{1}{\abs{\vf{r}_i - \vf{r}_j}}.
    \end{align}
    We thus get the total Hamiltonian
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i) + \sum_{i < j}^N v(\vf{r}_i, \vf{r}_j).
    \end{align}
    Working in a basis of $L$ single particle functions, $\{\ket{p}\}_{p =
    1}^L$. We define the reference Slater determinant as
    \begin{align}
        \ket{\Phi_0} &\equiv \ket{1, 2, \dots, N},
    \end{align}
    i.e., a tensorproduct of the $N$ first single particle functions, $\ket{i}$,
    of the system. We call these single particle functions \emph{occupied} as
    they are contained in the Slater determinant.  We will denote the occupied
    indices with $i, j, k, l, \dots \in \{1, \dots, N\}$, the \emph{virtual}
    states with $a, b, c, d, \dots \in \{N + 1, \dots, L\}$ and general indices
    with $p, q, r, s, \dots \in \{1, \dots, L\}$. In terms of sets of basis
    functions we can write this as
    \begin{align}
        \{\ket{p}\}_{p = 1}^L
        = \{\ket{i}\}_{i = 1}^N
        \cup
        \{\ket{a}\}_{a = N + 1}^L,
    \end{align}
    i.e., the general indexed states consists of both occupied and virtual
    states.

@@

    \subsection{The coupled cluster approximation}
        We approximate the true wavefunction, $\ket{\Psi}$, of the system by the
        coupled cluster wavefunction, $\ket{\Psi_{\text{CC}}}$, defined as
        \begin{align}
            \ket{\Psi_{\text{CC}}}
            &\equiv e^{T}\ket{\Phi_0}
            = \para{
                \sum_{i = 0}^n
                \frac{1}{n!}T^n
            }\ket{\Phi_0},
        \end{align}
        where the \emph{cluster operator}, $T$, is given by a sum of
        $p$-excitation operators
        \begin{align}
            T &= T_1 + T_2 + \dots + T_p \\
            &=
            \sum_{ia}t_i^a\acr{a}\ade{i}
            + \para{\frac{1}{2!}}^2\sum_{ijab}
            t_{ij}^{ab}\acr{a}\acr{b}\ade{i}\ade{j}
            + \dots,
        \end{align}
        where $t_{i\dots}^{a\dots}$
        We will be looking at the \emph{doubles} approximation, that is
        \begin{align}
            T = T_{\text{CCD}} \equiv T_2.
        \end{align}

    \subsection{Energy of the coupled cluster approximation}
        When we're going to compute the energy of a system using the coupled
        cluster approximation we would ideally want to find the expectation
        value of the energy using the coupled cluster wavefunction.
        \begin{align}
            E = \bcc H\kcc.
        \end{align}
        As it turns out, this is an uncomfortable way of finding the energy
        as $T \neq T^{\dagger}$. Instead we will define what we call the
        \emph{similarity transformed Hamiltonian}. We plug the coupled
        cluster wavefunction into the Schrödinger equation.
        \begin{align}
            H\kcc = E\kcc.
        \end{align}
        Next, we left multiply with the inverse of the cluster expansion,
        i.e.,
        \begin{align}
            e^{-T}H\kcc = e^{-T}E\kcc
            = E \kslat.
        \end{align}
        Projecting this equation on the reference state we get
        \begin{align}
            E = \bslat e^{-T}H\kcc
            = \bslat e^{-T}He^{T}\kslat,
        \end{align}
        where in the latter inner-product we have located the similarity
        transformed Hamiltonian defined by
        \begin{align}
            \bar{H} \equiv e^{-T}He^{T}.
            \label{eq:similarity_transformed_hamiltonian}
        \end{align}

        \subsubsection{Normal ordered Hamiltonian}
            To get to the coupled cluster equations we need to write the
            Hamiltonian on a normal ordered form. We use Wick's theorem on the
            Fermi vacuum\footnote{Fermi vacuum corresponds to the reference
            Slater.}. The one-body part of the Hamiltonian thus becomes
            \begin{align}
                h &= \sum_{pq}h^p_q\acr{p}\ade{q}
                = \sum_{pq}h^p_q\para{
                    \{\acr{p}\ade{q}\}
                    + \{
                        \wick{\c a_{p}^{\dagger} \c a_{q}}
                    \}
                }
                \\
                &= h_N + \sum_{pq}\varepsilon_i.
            \end{align}
            The two-body part yields
            \begin{align}
                v &= \sum_{pqrs}w^{pq}_{rs}\acr{p}\acr{q}\ade{s}\ade{r}
            \end{align}
            % TODO: Fill in the remainder. Preferably in the appendix.
            We can then write the Hamiltonian in terms of the reference energy
            and the normal ordered Hamiltonian.
            \begin{align}
                H = H_N + \bslat H\kslat.
            \end{align}
            The energy equation thus becomes
            \begin{align}
                E &= \bslat\bar{H}\kslat
                = E_{\text{ref}} + \bslat e^{-T}H_N e^T\kslat.
            \end{align}
            In this equation the unknowns are the cluster amplitudes

        By expanding the exponentials the similarity transformed Hamiltonian and
        recognizing the commutators we get the Baker-Campbell-Hausdorff
        expansion.
        \begin{align}
            \bar{H}
            &=
            H + \com{H}{T} + \half\com{\com{H}{T}}{T} + \dots.
        \end{align}

@@

>>>>

\newcommand{\ecc}{E_{\text{CC}}}
\newcommand{\eccd}{E_{\text{CCD}}}

@@

\section{Theory}
    In this project we will study a system of $N$ interacting electrons. We will
    be looking at a Hamiltonian consisting of a one-body and a two-body part.
    The one-body part is given by
    \begin{align}
        h(\vf{r}_i)
        &= -\half\nabla_i^2 + \half\omega^2 \vf{r}_i^2,
    \end{align}
    where we use natural units $\hbar = c = e = 1$ and set the mass to unity.
    The two-body part is the Coulomb interaction potential.
    \begin{align}
        v(\vf{r}_i, \vf{r}_j)
        &= \frac{1}{\abs{\vf{r}_i - \vf{r}_j}}.
    \end{align}
    We thus get the total Hamiltonian
    \begin{align}
        H &= h + W
        =
        \sum_{i = 1}^N h(\vf{r}_i) + \sum_{i < j}^N v(\vf{r}_i, \vf{r}_j),
    \end{align}
    where $h$ is the full one-body operator and $W$ the full two-body
    operator, i.e., over the entire system.  Working in a basis of $L$ single
    particle functions, $\{\ket{p}\}_{p = 1}^L$. We define the reference Slater
    determinant as
    \begin{align}
        \ket{\Phi_0} &\equiv \ket{1, 2, \dots, N},
    \end{align}
    i.e., a tensorproduct of the $N$ first single particle functions, $\ket{i}$,
    of the system. We call these single particle functions \emph{occupied} as
    they are contained in the Slater determinant.  We will denote the occupied
    indices with $i, j, k, l, \dots \in \{1, \dots, N\}$, the \emph{virtual}
    states with $a, b, c, d, \dots \in \{N + 1, \dots, L\}$ and general indices
    with $p, q, r, s, \dots \in \{1, \dots, L\}$. In terms of sets of basis
    functions we can write this as
    \begin{align}
        \{\ket{p}\}_{p = 1}^L
        = \{\ket{i}\}_{i = 1}^N
        \cup
        \{\ket{a}\}_{a = N + 1}^L,
    \end{align}
    i.e., the general indexed states consists of both occupied and virtual
    states. Note that the single particle functions are orthonormal, i.e.,
    \begin{align}
        \braket{p}{q} = \delta_{pq}.
    \end{align}
    We can construct other Slater determinants in this basis by exciting
    or de-exciting the reference determinant. A general excitation is labeled
    $\ket{\Phi_{ij\dots}^{ab\dots}}$ which means that we have removed the single
    particle functions with indices $i, j, \dots$ from the reference and added
    $a, b, \dots$. Note that
    \begin{align}
        \braket{\Phi^{ab\dots}_{ij\dots}}{\Phi_0} = 0,
    \end{align}
    for any excitation.

@@

    \subsection{The coupled cluster approximation}
        We approximate the true wavefunction, $\ket{\Psi}$, of the system by the
        coupled cluster wavefunction, $\ket{\Psi_{\text{CC}}}$, defined by
        \begin{align}
            \ket{\Psi_{\text{CC}}}
            &\equiv e^{T}\ket{\Phi_0}
            = \para{
                \sum_{i = 0}^n
                \frac{1}{n!}T^n
            }\ket{\Phi_0},
        \end{align}
        where the \emph{cluster operator}, $T$, is given by a sum of
        $p$-excitation operators labeled $T_p$. They consist of \emph{cluster
        amplitudes}, $t_{i\dots}^{a\dots}$, and creation and annihilation
        operators.
        \begin{align}
            T &= T_1 + T_2 + \dots + T_p \\
            &=
            \sum_{ia}t_i^a\acr{a}\ade{i}
            + \para{\frac{1}{2!}}^2\sum_{ijab}
            t_{ij}^{ab}\acr{a}\acr{b}\ade{i}\ade{j}
            + \dots.
        \end{align}
        In the doubles approximation we limit the cluster operator to
        \begin{align}
            T \equiv T_2
            = \frac{1}{4}\sum_{ijab}t_{ij}^{ab}\acr{a}\acr{b}\ade{j}\ade{i}.
            \label{eq:T_2}
        \end{align}
        The first part of the coupled cluster method consists of constructing
        the cluster amplitudes using the \emph{amplitude equations}. After we
        have found the amplitudes we can compute the energy.

    \subsection{Energy of the coupled cluster approximation}
        When we're going to compute the energy of a system using the coupled
        cluster approximation we would ideally want to find the expectation
        value of the energy using the coupled cluster wavefunction.
        \begin{align}
            \ecc = \bcc H\kcc.
        \end{align}
        As it turns out, this is an uncomfortable way of finding the energy
        as $T \neq T^{\dagger}$. Instead we will define what we call the
        \emph{similarity transformed Hamiltonian}. We plug the coupled
        cluster wavefunction into the Schrödinger equation.
        \begin{align}
            H\kcc = \ecc\kcc.
        \end{align}
        Next, we left multiply with the inverse of the cluster expansion,
        i.e.,
        \begin{align}
            e^{-T}H\kcc = e^{-T}\ecc\kcc
            = \ecc \kslat.
        \end{align}
        Projecting this equation on the reference state we get
        \begin{align}
            \ecc = \bslat e^{-T}H\kcc
            = \bslat e^{-T}He^{T}\kslat,
        \end{align}
        where in the latter inner-product we have located the similarity
        transformed Hamiltonian defined by
        \begin{align}
            \bar{H} \equiv e^{-T}He^{T}.
            \label{eq:similarity_transformed_hamiltonian}
        \end{align}

        To simplify the energy equation and the amplitude equations we use the
        normal ordered Hamiltonian.
        \begin{align}
            H = H_N + \bslat H\kslat.
        \end{align}
        The energy equation thus becomes
        \begin{align}
            \ecc &= \bslat\bar{H}\kslat
            = E_{\text{ref}} + \bslat e^{-T}H_N e^T\kslat,
        \end{align}
        where the reference energy is given by
        \begin{align}
            E_{\text{ref}} = \bslat H\kslat.
        \end{align}
        We now define the normal ordered similiarity transformed Hamiltonian as
        \begin{align}
            \bar{H}_N \equiv e^{-T}H_N e^T.
        \end{align}
        By expanding the exponentials of this Hamiltonian and recognizing the
        commutators we get the Baker-Campbell-Hausdorff expansion.
        \begin{align}
            \bar{H}_N
            &=
            H_N + \com{H_N}{T} + \frac{1}{2!}\com{\com{H_N}{T}}{T} + \dots.
        \end{align}
        From the connected cluster theorem we know that the only nonzero terms
        in the Baker-Campbell-Hausdorff expansion will be the terms where the
        normal ordered Hamiltonian has at least one contraction\footnote{In the
        Wick's theorem sense.} with every cluster operator on its right. This
        lets us write the expansion as
        \begin{align}
            \bar{H}_N
            &=
            H_N + \para{H_N T}_c + \frac{1}{2!}\para{H_N T^2}_c + \dots,
            \label{eq:sim_norm_hamiltionian_expansion}
        \end{align}
        where the subscript $c$ signifies that only contributions where at least
        one contraction between $H_N$ and $T$ has been performed will be
        included.

        \subsubsection{Coupled cluster doubles energy equation}
            Using the doubles approximation with the cluster operator $T_2$
            defined in \autoref{eq:T_2} the energy equation becomes
            \begin{align}
                \eccd
                &=
                E_{\text{ref}}
                + \bslat e^{-T_2}H_N e^{T_2}\kslat.
            \end{align}
            As the doubles cluster operator doubly excites the reference and
            using the expansion in \autoref{eq:sim_norm_hamiltionian_expansion}
            we see that we can write the energy equation as
            \begin{align}
                \eccd
                &=
                E_{\text{ref}}
                + \bslat H_N(1 + T_2)\kslat,
            \end{align}
            as the Hamiltonian is only able to de-excite one pair of single
            particle functions.

@@

\appendix
\section{The normal ordered Hamiltonian}
    When constructing the normal ordered Hamiltonian we use Wick's theorem to
    write the one-body, $h$, and the two-body, $W$, operators onto a normal
    ordered form. Specifically we define the normal ordered form in terms of
    Fermi vacuum. That is, an operator on normal ordered form destroys the
    reference Slater determinant.
    \begin{align}
        h &= \sum_{pq}h^p_q\acr{p}\ade{q}
        = \sum_{pq}h^p_q\para{
            \{\acr{p}\ade{q}\}
            + \{
                \wick{\c a_{p}^{\dagger} \c a_{q}}
            \}
        }
        \\
        &= \sum_{pq}h^p_q\{
            \acr{p}\ade{q}
        \}
        + \sum_{pq}h^p_q\delta_{p \in i}\delta_{pq}
        \\
        &= h_N + \sum_{i}h^i_i,
    \end{align}
    where we have used $\delta_{p \in i}$ to mean that $p$ must be an occupied
    index.
    \cite{crawford2007introduction}
    % TODO: Fill in the rest of this expansion
